Animacy Detection in Stories

### 2014-12-12 paper/paper.tex
kappa scores and new results

\usepackage{blindtext}
\bibliographystyle{plain}

@@

For almost all species in the world, the capacity to distinguish
animate objects from inanimate objects is essential to their
survival. Those objects could be prey, for example, or predators or
mates. The fundamental nature of the distinction between animate and
inanimate is reflected in the fact that the division is acquired very early in life.
\cite{opfer:02}. Children of less than 6 months old are well able to
make the division. Moreover, recent brain research shows that the distinction
appears in the organization of the brain (e.g.~\cite{gao:12}). For some
researchers this provides evidence for the idea that the division
between animate and inanimate is an innate part of how we see the
world.
%hoe sluit dit aan op de eerdere alinea? lijkt alsof je jezelf tegenspreekt namelijk: als het een standpunt is, waarom zien we dat dan niet in de hersens?%
A common view supported by philosophers such as Daniel Dennett is that
animacy and aliveness are better to be treated as epistemological
stances rather than fixed states in the world \cite{dennett:96}: not
ineffable qualia but behavioral capacity defines our stance towards
objects. Depending on whether people \textit{think} that an object is
animate, they utilize different cognitive strategies to explain and
predict the actions of those objects. Th following opening of a
well-known story will make this point more clear:

@@

\cite{moore:13} present an interesting approach that combines a number
of animacy classifiers in a voting scheme and aims at an interpretable
and correctable model of animacy classification. They combine a
variety of classifiers such as the WordNet-based approach of
\cite{evans:00}, dictionary sources and systems for named entity
recognition.

% commentary on above approaches
The approaches mentioned above present us with a number of problems. First, almost all of them heavily rely on costly,
linguistically informed features derived from lexical-semantic
databases or syntactic parsing. For most languages in the world,
however, we cannot rely on these resources, either because they do not exist, or because %syntactic parsers have too low results%. 
Second, animacy detection is often seen as a useful feature for a range of natural language
processing techniques, such as anaphora resolution and syntactic
parsing. The mutual dependence of these techniques and animacy
detection, however, leads to a chicken and egg situation. Another
problem with the approaches above is that they are lemma-based, which
means that the models are generally insensitive to different usages of
a particular word. A pancake is typically not an animate entity, but,
as we have seen, whether it is animate or not may depend on the story in which it appears, and even within the story the animacy may change. We therefore need to
develop a dynamic, data driven system that is token based and does not
rely too much on static and fixed world knowledge.

@@

To develop this dynamic, data driven system we used  a corpus of Dutch folktales. The corpus consists of 74 Dutch stories from the collection
\textit{Volkssprookjes uit Nederland en Vlaanderen} compiled by
\cite{sinninghe:78}. The collection is made up of Dutch retellings of popular stories, including such tales as \textit{The Bremen
  Town Musicians} (ATU 130) and \textit{The Table, the Ass, and the
  Stick } (ATU 563), as well as lesser-known stories such as
\textit{The Singing Bone} (ATU 780) and \textit{Cock, Hen, Duck, Pin, and Needle on a Journey} (AT 0210). This last story is again a clear example of otherwise inanimate objects being animate in a folk story, as is concerns the adventures of several household items, such as a pin, a hackle, an egg, and a whetstone. A digital version of the
collection is available in the Dutch Folktale Database from the
Meertens Institute (corpus SINVSUNV.20E).\footnote{See
\url{http://www.verhalenbank.nl}} Using a single collection for our corpus presents us with a helpful homogeneity with regard to the editor, length and language use. On the other hand, the collection displays diversity as well in the choice of the stories, which entail fairy tales, legends, and sagas.
%getallen nakijken%
%moet je de labels hier al wel noemen? want dat lijkt me eerder een resultaat dan een pre-processing stap toch?%

The corpus consists of 74504 words, from 5549 unique words. All words or tokens were assigned a binary label `animate' or `inanimate', where all words that were not animate (e.g. determiners, adjectives, etc.) were labelled as inanimate. This resulted in 11542 animate tokens of 743 types, and 62926 inanimate tokens from 5011 unique inanimate words. A selection of words, such as the abovementioned pancake and egg, was labeled as both animate and inanimate, depending on the context. The distribution of the nouns is striking: of the animate tokens, 4627 (40,1\%) are nound, while only 6878 of the inanimate tokens (11 \%) are nouns. This shows that being a noun is already somewhat of an indication for animacy.

% voorbeeld met gloss 4 regels (NL, ani/inani/Engels/vrij vertaling)

@@

In real world applications, an animacy detection system will most
likely be faced with completely new texts instead of single words. It
is therefore important to construct a training and test procedure in
such a way that it mimics this situation as closely as possible. If we
would, for example, make a random split of 80\% of the data for training and
20\% for testing on the word level, we run the risk of mixing training
data with test data, thereby making it too easy for a system to rely on words
it has seen from the same text. \cite{bowman:12} fall into this trap
by making a random split in their data on the sentence level. It is
very likely that in their setup sentences from the same document are
present in both the training data and the test data, making their
evaluation unrealistic. To circumvent this problem, we split at the story level, using 80\% of the stories (n=59) as training material and the remaining 15 stories as test.
%dat is de innovatie right? dat het niet gaat om woord of zin maar om de hele tekst%

@@

In table \ref{tab:results-all} we present the results for all nine
models on the complete data set. For each model we report the
precision, recall and $F1$-score for the animate words and the
inanimate words.

\begin{table}
\centering
\begin{tabular}{llrrrr}
\toprule
features                                                 &           & precision & recall & $F1$ & avg $F1$              \\ \midrule
\multirow{2}{*}{word}                                    & animate   & 0.93      & 0.80   & 0.86 & \multirow{2}{*}{0.96} \\
                                                         & inanimate & 0.96      & 0.99   & 0.98 &                       \\
\multirow{2}{*}{word + PoS}                              & animate   & 0.92      & 0.88   & 0.90 & \multirow{2}{*}{0.97} \\
                                                         & inanimate & 0.98      & 0.99   & 0.98 &                       \\
\multirow{2}{*}{word + PoS + lemma}                      & animate   & 0.93      & 0.89   & 0.91 & \multirow{2}{*}{0.97} \\
                                                         & inanimate & 0.98      & 0.99   & 0.98 &                       \\
\multirow{2}{*}{word + PoS + lemma + dep}                & animate   & 0.93      & 0.89   & 0.91 & \multirow{2}{*}{0.97} \\
                                                         & inanimate & 0.98      & 0.99   & 0.98 &                       \\
\multirow{2}{*}{word + embeddings}                       & animate   & 0.92      & 0.93   & 0.92 & \multirow{2}{*}{0.98} \\
                                                         & inanimate & 0.99      & 0.99   & 0.99 &                       \\
\rowcolor{Gray}                                          & animate   & 0.92      & 0.93   & 0.93 &                       \\
\rowcolor{Gray}\multirow{-2}{*}{word + PoS + embeddings} & inanimate & 0.99      & 0.99   & 0.99 & \multirow{-2}{*}{0.98}\\
\multirow{2}{*}{word + PoS + lemma + embeddings}         & animate   & 0.93      & 0.93   & 0.93 & \multirow{2}{*}{0.98} \\
                                                         & inanimate & 0.99      & 0.99   & 0.99 &                       \\
\multirow{2}{*}{word + PoS + lemma + dep + embeddings}   & animate   & 0.92      & 0.93   & 0.93 & \multirow{2}{*}{0.98} \\
                                                         & inanimate & 0.99      & 0.99   & 0.99 &                       \\
\multirow{2}{*}{word embeddings}                         & animate   & 0.90      & 0.92   & 0.91 & \multirow{2}{*}{0.97} \\
                                                         & inanimate & 0.99      & 0.98   & 0.98 &                       \\

\bottomrule
\end{tabular}
\caption{Precision, recall and $F1$ scores for animate and inanimate classes per feature settings for all words.}
\label{tab:results-all}
\end{table}

All models perform well on classifying inanimate
words. However, since this is the majority class, it is far more
interesting to compare the performance of the models on the animate
instances. It is interesting to observe that the `simple' $n$-gram
word model already performs rather well. Adding more features, like
Part-of-Speech, lemmata, etcetera, only has a positive impact on the
recall of the model, while leaving the precision untouched. As can be
observed from the table, employing the rather expensive dependency features show
barely any improvement. Unexpectedly, the model that only uses word embedding
features is one of the best performing models. This was slightly
unexpected, since this is the model in which features are context
insensitive. However, the model that combines the word embeddings with
the word $n$-gram features outperforms the other linguistically
informed models that make use of lemmata, part-of-speech tags or
dependency information. The best performance is achieved by the model
that combines the word features, part-of-speech tags and the word
embeddings with an $F1$-score of 0.93 for animate words and 0.99 on
inanimate words. Adding more features does not result in any more
performance gain.

Table \ref{tab:results-noun} presents the results for all nouns and
names in the data set. The best performance is again achieved by the
model that combines the word features with the Part-of-Speech tags and
word embeddings, resulting in an $F1$-score of 0.92 for animate
instances and 0.95 for inanimate instances.

\begin{table}
\centering
\begin{tabular}{llrrrr}
\toprule
features                                               &           & precision & recall & $F1$ & avg $F1$              \\ \midrule
\multirow{2}{*}{word}                                  & animate   & 0.96      & 0.64   & 0.77 & \multirow{2}{*}{0.84} \\
                                                       & inanimate & 0.81      & 0.98   & 0.89 &                       \\
\multirow{2}{*}{word + PoS}                            & animate   & 0.92      & 0.80   & 0.86 & \multirow{2}{*}{0.90} \\
                                                       & inanimate & 0.89      & 0.95   & 0.92 &                       \\
\multirow{2}{*}{word + PoS + lemma}                    & animate   & 0.93      & 0.82   & 0.87 & \multirow{2}{*}{0.90} \\
                                                       & inanimate & 0.89      & 0.96   & 0.92 &                       \\
\multirow{2}{*}{word + PoS + lemma + dep}              & animate   & 0.93      & 0.82   & 0.87 & \multirow{2}{*}{0.91} \\
                                                       & inanimate & 0.90      & 0.96   & 0.93 &                       \\
\multirow{2}{*}{word + embeddings}                     & animate   & 0.92      & 0.91   & 0.92 & \multirow{2}{*}{0.94} \\
                                                       & inanimate & 0.95      & 0.95   & 0.95 &                       \\
\rowcolor{Gray}                                        & animate   & 0.92      & 0.92   & 0.92 &                       \\
\rowcolor{Gray}\multirow{-2}{*}{word + PoS + embeddings} & inanimate & 0.95    & 0.95   & 0.95 & \multirow{-2}{*}{0.94}\\
\multirow{2}{*}{word + PoS + lemma + embeddings}       & animate   & 0.92      & 0.91   & 0.92 & \multirow{2}{*}{0.94} \\
                                                       & inanimate & 0.95      & 0.95   & 0.95 &                       \\
\multirow{2}{*}{word + PoS + lemma + dep + embeddings} & animate   & 0.92      & 0.91   & 0.92 & \multirow{2}{*}{0.94} \\
                                                       & inanimate & 0.94      & 0.95   & 0.95 &                       \\
\multirow{2}{*}{word embeddings}                       & animate   & 0.88      & 0.90   & 0.89 & \multirow{2}{*}{0.92} \\
                                                       & inanimate & 0.94      & 0.92   & 0.93 &                       \\

\bottomrule
\end{tabular}
\caption{Precision, recall and $F1$ scores for animate and inanimate classes
  per feature settings for all words tagged as noun.}
\label{tab:results-noun}
\end{table}

>>>>

\usepackage{gb4e}
\bibliographystyle{plain}

@@

For almost all species in the world, the capacity to distinguish
animate objects from inanimate objects is essential to their
survival. Those objects could be prey, for example, or predators or
mates. The fundamental nature of the distinction between animate and
inanimate is reflected in the fact that the division is acquired very
early in life~\cite{opfer:02}. Children of less than 6 months old
are well able to make the division. Moreover, recent brain research
shows that the distinction appears in the organization of the brain
(e.g.~\cite{gao:12}). For some researchers this provides evidence for
the idea that the division between animate and inanimate is an innate
part of how we see the world.

A common view supported by philosophers such as Daniel Dennett is that
animacy and aliveness are better to be treated as epistemological
stances rather than fixed states in the world \cite{dennett:96}: not
ineffable qualia but behavioral capacity defines our stance towards
objects. Depending on whether people \textit{think} that an object is
animate, they utilize different cognitive strategies to explain and
predict the actions of those objects. Th following opening of a
well-known story will make this point more clear:

@@

\cite{moore:13} present an approach that combines a number of animacy
classifiers in a voting scheme and aims at an interpretable and
correctable model of animacy classification. They combine a variety of
classifiers such as the WordNet-based approach of \cite{evans:00},
dictionary sources and systems for named entity recognition.

% commentary on above approaches
The approaches mentioned above present us with a number of
problems. First, almost all of them heavily rely on costly,
linguistically informed features derived from lexical-semantic
databases or syntactic parsing. For most languages in the world,
however, we cannot rely on these resources, either because they do not
exist, or because their performance is insufficient.  Second, animacy
detection is often seen as a useful feature for a range of natural
language processing techniques, such as anaphora resolution and
syntactic parsing. The mutual dependence of these techniques and
animacy detection, however, leads to a chicken and egg
situation. Another problem with the approaches above is that they are
lemma-based, which means that the models are generally insensitive to
different usages of a particular word. A pancake is typically not an
animate entity, but, as we have seen, whether it is animate or not may
depend on the story in which it appears, and even within the story the
animacy may change. We therefore need to develop a dynamic, data
driven system that is token based and does not rely too much on static
and fixed world knowledge.

@@

To develop this dynamic, data driven system we used a corpus of Dutch
folktales. The corpus consists of 74 Dutch stories from the collection
\textit{Volkssprookjes uit Nederland en Vlaanderen} compiled by
\cite{sinninghe:78}. The collection is made up of Dutch retellings of
popular stories, including such tales as \textit{The Bremen Town
  Musicians} (ATU 130) and \textit{The Table, the Ass, and the Stick }
(ATU 563), as well as lesser-known stories such as \textit{The Singing
  Bone} (ATU 780) and \textit{Cock, Hen, Duck, Pin, and Needle on a
  Journey} (ATU 210). This last story is again a clear example of
otherwise inanimate objects being animate in a folk story, as it
concerns the adventures of several household items, such as a
\textit{pin}, a \textit{hackle}, an \textit{egg}, and a
\textit{whetstone}. A digital version of the collection is available
in the Dutch Folktale Database from the Meertens Institute (corpus
SINVSUNV.20E).\footnote{See \url{http://www.verhalenbank.nl}} Using a
single collection for our corpus presents us with a helpful
homogeneity with regard to the editor, length and language use. On the
other hand, the collection displays diversity as well in the choice of
the stories, which entail fairy tales, legends, and sagas.
%getallen nakijken%
%moet je de labels hier al wel noemen? want dat lijkt me eerder een resultaat dan een pre-processing stap toch?%

The data were labeled by two annotators. We computed an
inter-annotator agreement score (Cohen's Kappa) of $K=0.95$ on the
basis of five stories that were annotated by both annotators. The
following sentence provides an example annotation. All unlabeled words
are implicitly considered to be inanimate.
\begin{exe}
\ex
\gll Jij smid, jij bent de sterkste; hou je vast aan de bovenste
takken, en dan ga jij kleermaker aan zijn benen hangen en zo gaan we maar door\\
    \textsc{anim} \textsc{anim} \textsc{anim} {} {} {} {} \textsc{anim} {} {} {} {} {} {} {} {} \textsc{anim} \textsc{anim}
    {} \textsc{anim} {} {} {} {} {} \textsc{anim} {} {} \\
\trans `You, blacksmith, you are the strongest; hold on to the upper
branches and then, you, tailor, will grab his legs and so we go on\ldots'
\end{exe}
The corpus consists of 74504 words, from 5549 unique words. All word
tokens were assigned a binary label `animate' or `inanimate', where
all words that were not animate (e.g. determiners, adjectives, etc.)
were labelled as inanimate. This resulted in 11542 animate tokens of
743 types, and 62926 inanimate tokens from 5011 unique inanimate
words. A selection of words, such as the abovementioned
\textit{pancake} and \textit{egg}, was labeled as both animate and
inanimate, depending on the context. The distribution of the nouns is
telling: of the animate tokens, 4627 (40,1\%) are nound, while only
6878 of the inanimate tokens (11 \%) are nouns. This shows that being
a noun is already somewhat of an indication for animacy.

@@

In real world applications, an animacy detection system will most
likely be faced with completely new texts instead of single words. It
is therefore important to construct a training and test procedure in
such a way that it mimics this situation as closely as possible. If we
would, for example, make a random split of 80\% of the data for
training and 20\% for testing on the word level, we run the risk of
mixing training data with test data, thereby making it too easy for a
system to rely on words it has seen from the same
text. \cite{bowman:12} fall into this trap by making a random split in
their data on the sentence level. It is very likely that in their
setup sentences from the same document are present in both the
training data and the test data, making their evaluation
unrealistic. To circumvent this problem, we split the data at the
story level. We make use of 10-fold cross-validation, where we
randomly select 10\% of the stories for testing and train on the
remaining 90\%.

@@

In table \ref{tab:results-all} we present the results for all nine
models on the complete data set. For each model we report the
precision, recall, $F1$-score and Area Under the Precision-Recall
Curve (AUC) for the animate words and the
inanimate words.

\begin{table}
\centering
\begin{tabular}{llrrrr}
\toprule
features & class & precision &  recall & $F1$ &  AUC \\
\midrule
\multirow{2}{*}{embeddings}                            & inanimate     &       0.98 &    0.99 &    0.98 & \multirow{2}{*}{0.93} \\
                                                       & animate     &       0.92 &    0.89 &    0.91 &  \\
\multirow{2}{*}{word}                                  & inanimate     &       0.96 &    0.99 &    0.98 & \multirow{2}{*}{0.94} \\
                                                       & animate     &       0.94 &    0.78 &    0.85 &  \\
\multirow{2}{*}{word + embeddings}                     & inanimate     &       0.98 &    0.99 &    0.98 & \multirow{2}{*}{0.96} \\
                                                       & animate     &       0.93 &    0.90 &    0.91 &  \\
\multirow{2}{*}{word + PoS}                            & inanimate     &       0.97 &    0.99 &    0.98 & \multirow{2}{*}{0.96} \\
                                                       & animate     &       0.94 &    0.86 &    0.90 &  \\
\rowcolor{Gray}                                        & inanimate     &       0.98 &    0.99 &    0.99 &  \\
\rowcolor{Gray}\multirow{-2}{*}{word + PoS + embeddings}   & animate     &       0.94 &    0.92 &    0.93 & \multirow{-2}{*}{0.97} \\
\multirow{2}{*}{word + PoS + lemma}                    & inanimate    &       0.98 &    0.99 &    0.98 & \multirow{2}{*}{0.96} \\
                                                       & animate     &       0.94 &    0.87 &    0.90 &  \\
\multirow{2}{*}{word + PoS + lemma + embeddings}       & inanimate     &       0.98 &    0.99 &    0.99 & \multirow{2}{*}{0.97} \\
                                                       & animate     &       0.94 &    0.92 &    0.93 &  \\
\multirow{2}{*}{word + PoS + lemma + dep}              & inanimate     &       0.98 &    0.99 &    0.98 & \multirow{2}{*}{0.96} \\
                                                       & animate     &       0.94 &    0.88 &    0.91 &  \\
\multirow{2}{*}{word + PoS + lemma + dep + embeddings} & inanimate     &       0.98 &    0.99 &    0.99 & \multirow{2}{*}{0.97} \\
                                                       & animate     &       0.94 &    0.92 &    0.93 &  \\
\bottomrule
\end{tabular}
\caption{Precision, recall, $F1$ and Area Under the Curve (AUC) scores for animate and inanimate classes per feature setting for all words.}
\label{tab:results-all}
\end{table}

All models perform well on classifying inanimate words. However, since
this is the majority class, it is of course far more interesting to
compare the performance of the models on the animate instances. It is
interesting to observe that the `simple' $n$-gram word model already
performs rather well. Adding more features, like Part-of-Speech,
lemmata, etcetera, only has a positive impact on the recall of the
model, while leaving the precision untouched. As can be observed from
the table, employing the rather expensive dependency features shows
barely any improvement.

The model that only uses word embedding features is one of the best
performing models. This is a context insensitive model that operates
on the level of the vocabulary, which means that it will predict the
same outcome for each token of a particular word type. The high
precision and high recall show us that this model has acquired
knowledge about which words \textit{typically} group with animate
words and which with inanimate words. However, the models that combines
the word embeddings with the context sensitive features, such as word
$n$-grams or Part-of-Speech tags, outperform the context insensitive
model.

The best performance is
achieved by the model that combines the word features, part-of-speech
tags and the word embeddings with an $F1$-score of 0.93 for animate
words and 0.99 on inanimate words. Adding more features does not
result in any more performance gain.

Table \ref{tab:results-noun} presents the results for all nouns and
names in the data set. The best performance is again achieved by the
model that combines the word features with the Part-of-Speech tags and
word embeddings, resulting in an $F1$-score of 0.92 for animate
instances and 0.94 for inanimate instances.

\begin{table}
\centering
\begin{tabular}{llrrrr}
\toprule
features & class &  precision &  recall & $F1$ &  AUC \\
\midrule
\multirow{2}{*}{embeddings}                            & inanimate     &       0.89 &    0.94 &    0.92 & \multirow{2}{*}{0.92} \\
                                                       & animate       &       0.91 &    0.85 &    0.88 &  \\
\multirow{2}{*}{word}                                  & inanimate     &       0.78 &    0.98 &    0.87 & \multirow{2}{*}{0.91} \\
                                                       & animate       &       0.96 &    0.60 &    0.74 &  \\
\multirow{2}{*}{word + embeddings}                     & inanimate     &       0.90 &    0.96 &    0.93 & \multirow{2}{*}{0.96} \\
                                                       & animate       &       0.94 &    0.86 &    0.89 &  \\
\multirow{2}{*}{word + pos}                            & inanimate     &       0.86 &    0.96 &    0.90 & \multirow{2}{*}{0.94} \\
                                                       & animate       &       0.92 &    0.78 &    0.84 &  \\
\rowcolor{Gray}                                        & inanimate     &       0.93 &    0.96 &    0.94 &  \\
\rowcolor{Gray}\multirow{-2}{*}{word + PoS + embeddings} & animate       &       0.94 &    0.89 &    0.92 &\multirow{-2}{*}{0.97}  \\
\multirow{2}{*}{word + PoS + lemma}                    & inanimate     &       0.87 &    0.96 &    0.91 & \multirow{2}{*}{0.95} \\
                                                       & animate       &       0.94 &    0.80 &    0.86 &  \\
\multirow{2}{*}{word + PoS + lemma + embeddings}       & inanimate     &       0.93 &    0.96 &    0.94 & \multirow{2}{*}{0.97} \\
                                                       & animate       &       0.94 &    0.89 &    0.92 &  \\
\multirow{2}{*}{word + PoS + lemma + dep}              & inanimate     &       0.87 &    0.96 &    0.91 & \multirow{2}{*}{0.95} \\
                                                       & animate       &       0.93 &    0.80 &    0.86 &  \\
\multirow{2}{*}{word + PoS + lemma + dep + embeddings} & inanimate     &       0.93 &    0.96 &    0.94 & \multirow{2}{*}{0.97} \\
                                                       & animate       &       0.94 &    0.90 &    0.92 &  \\
\bottomrule
\end{tabular}
\caption{Precision, recall, $F1$ and Area Under the Curve (AUC) scores for animate and inanimate classes
  per feature settings for all words tagged as noun.}
\label{tab:results-noun}
\end{table}
